---
title: "Effects of weather events on population health & the economy between 1950 and 2011"
author: "Rinnette Ramdhanie"
date: "19 September 2019"
output: 
        html_document: 
                keep_md: true
                number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Synopsis

This report aims to show how severe weather events recorded  during the period 1950 to November 2011 have affected population health and the economy in the United States.  This information is useful so that resources can be prioritised for the different types of events.  Population Health was measured by injuries and fatalities while economic consequences was measured by using cost of damages to property and crops.  It was found that tornadoes, thunderstorm winds, excessive heat, floods and lightning caused the most fatalities and injuries.  Floods, including flash floods, hurricanes/typhoons, hail and thunderstorm winds had the largest economic consequences.


# Loading and Processing the Raw Data
The U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database tracks characteristics of major storms and weather events in the United States, including when and where they occur.  

## Loading the data
The data was in the form of a comma-separated-value file compressed via the bzip2 algorithm.  It was loaded directly from the zipped file.

```{r loadData}
        library(dplyr)
        stormdata <- tbl_df(read.csv("repdata_data_StormData.csv.bz2"))
```

## Processing the data
The number of rows, columns and structure of the dataset were checked, as well as the first and last few rows.

```{r checkData}
        dim(stormdata)
        str(stormdata)
        head(stormdata)
        tail(stormdata)
```

The data seems to have been loaded correctly. 

The documentation provided with the data describes 48 different event types. The number of unique event types in the database was checked.

```{r numEvents}
        length(unique(stormdata$EVTYPE))
```

985 uniques values in the EVTYPE column shows a huge variation in how information was entered for this variable.  There seemes to be incorrect spelling, variations of the same name, combinations of several names and so on.  

It was decided that some of these event type names should be replaced with what is in the documentation, but only those that might affect the results of the analysis.  A description of what was done, as well as justification of the changes made are provided in the sections below.


### Processing the data to determine how events affect Population Health
The FATALITIES and INJURIES variables were used to show how weather events affect population health.  Since there were so many more event type names in the dataset compared to what was in the documentation provided, some checks were done to see how these 'other' names contribute to the numbers of fatalities and injuries.


#### Justification for replacing event names - checking injuries and fatalities
A vector with the 48 event types described in the documentation was created.  The raw data was then filtered to obtain those rows with event types that are NOT included in this vector. The proportions of FATALITIES and INJURIES attributed to these invalid event types were then calculated.


```{r percentInjFat}
        # Create vector of valid event types
                event_types <- c("ASTRONOMICAL LOW TIDE", "AVALANCHE", "BLIZZARD", "COASTAL FLOOD", "COLD/WIND CHILL", "DEBRIS FLOW", "DENSE FOG", "DENSE SMOKE", "DROUGHT", "DUST DEVIL", "DUST STORM", "EXCESSIVE HEAT", "EXTREME COLD/WIND CHILL", "FLASH FLOOD", "FLOOD", "FREEZING FOG", "FROST/FREEZE", "FUNNEL CLOUD", "HAIL", "HEAT", "HEAVY RAIN", "HEAVY SNOW", "HIGH SURF", "HIGH WIND", "HURRICANE/TYPHOON", "ICE STORM", "LAKESHORE FLOOD", "LAKE-EFFECT SNOW", "LIGHTNING", "MARINE HAIL", "MARINE HIGH WIND", "MARINE STRONG WIND", "MARINE THUNDERSTORM WIND", "RIP CURRENT", "SEICHE", "SLEET", "STORM TIDE", "STRONG WIND", "THUNDERSTORM WIND", "TORNADO", "TROPICAL DEPRESSION", "TROPICAL STORM", "TSUNAMI", "VOLCANIC ASH", "WATERSPOUT", "WILDFIRE", "WINTER STORM", "WINTER WEATHER")

        # Filter invalid event types from the data
                invEvents <- filter(stormdata, !(EVTYPE %in% event_types))
        
        # Calculate how much invalid event types contribute to the total fatalities and injuries
                sum(invEvents$FATALITIES)/sum(stormdata$FATALITIES)
                sum(invEvents$INJURIES)/sum(stormdata$INJURIES)

```

About 13% of fatalities and 9% of injuries can be attributed to events with names not in the documentation.  These amounts are large enough to justify replacing some of these names with the documented names.  


#### Description of the selection process for replacing event names
It may not be necessary to replace every invalid event type.  Aome analysis was done to determine which ones should be replaced.  

The INJURIES variable was summed for each invalid event name.

```{r checkInjuries}

        # Filter rows with invalid event names and injuries are not 0
                storm_i <- filter(stormdata, !(EVTYPE %in% event_types) & INJURIES != 0)

        # Sum injuries for each event type, and arrange in descending order
                isum <- storm_i %>%
                        group_by(EVTYPE) %>%
                        summarise (inj_sum = sum(INJURIES)) %>%
                        arrange(desc(inj_sum))

        # Summary of injuries
                summary(isum$inj_sum)
                head(isum, 20)
   
```

Sum of injuries for invalid event names which were less than 100 only contribute to about 1% of the total injuries which is very small. 

```{r}
        lessthan100_i <- filter(isum, !(EVTYPE %in% event_types) & inj_sum < 100)
        sum(lessthan100_i$inj_sum)/sum(stormdata$INJURIES)
```

A decision was therefore made to replace those event names with numbers of injuries greater than 100.

```{r selectNamesInj}
        
        # Filter rows where injuries are over 100
                morethan100_i <- filter(isum, !(EVTYPE %in% event_types) & inj_sum > 100)  
                
        # Select unique event names with over 100 injuries
                unique(morethan100_i$EVTYPE)
```

The raw dataset was copied to **sd_edited** and event names were replaced in this copied dataset.

```{r replaceInj}

        # Raw data copied
                sd_edited <- stormdata
        
        # Ensure all values in the EVTYPE variable are in upper case
                sd_edited$EVTYPE <- toupper(sd_edited$EVTYPE)
        
        # Replace event names
                sd_edited$EVTYPE <- gsub("^TSTM WIND$|^THUNDERSTORM WINDS$", "THUNDERSTORM WIND", 
                                         sd_edited$EVTYPE)
                sd_edited$EVTYPE <- gsub("^FOG$", "DENSE FOG", sd_edited$EVTYPE)
                sd_edited$EVTYPE <- gsub("^WILD/FOREST FIRE$|^WILD FIRES$", "WILDFIRE", 
                                         sd_edited$EVTYPE)
                sd_edited$EVTYPE <- gsub("^HEAT WAVE$|^EXTREME HEAT$", "EXCESSIVE HEAT", 
                                         sd_edited$EVTYPE)
                sd_edited$EVTYPE <- gsub("^HIGH WINDS$|^STRONG WINDS$|^STRONG WIND$", "HIGH WIND", 
                                         sd_edited$EVTYPE)
                sd_edited$EVTYPE <- gsub("^RIP CURRENTS$", "RIP CURRENT", sd_edited$EVTYPE)
                sd_edited$EVTYPE <- gsub("^EXTREME COLD$", "EXTREME COLD/WIND CHILL", 
                                         sd_edited$EVTYPE)
                sd_edited$EVTYPE <- gsub("^GLAZE$", "FROST/FREEZE", sd_edited$EVTYPE)
                sd_edited$EVTYPE <- gsub("^ICE$", "ICE STORM", sd_edited$EVTYPE)

```

A similar check was done for FATALITIES.

```{r checkFatalities}

        # Filter rows with invalid event names and fatalities are not 0
                storm_f <- filter(sd_edited, !(EVTYPE %in% event_types) & FATALITIES != 0)

        # Sum fatalities for each event type, and arrange in descending order
                fsum <- storm_f %>%
                        group_by(EVTYPE) %>%
                        summarise (fat_sum = sum(FATALITIES)) %>%
                        arrange(desc(fat_sum))

        # Calculate contribution to total fatalities
                head(fsum)
                sum(fsum$fat_sum)/sum(sd_edited$INJURIES)
```

No further event names were replaced as the sum of fatalities for invalid names contribute to less than 1% of the total fatalities.


### Processing the data to show how events affect the economy

The following variables were used to determine how events affect the economy:

* PROPDMG - contains figures representing estimates of the damage done to property
* PROPDMGEXP - an alpha character that signifies the magnitude of the amounts in PROPDMG, ie.
        + H - hundreds
        + K - thousands
        + M - millions
        + B - billions
* CROPDMG - contains figures representing estimates of the damage done to crops
* CROPDMGEXP - an alpha character that signifies the magnitude of the amounts in CROPDMG.  The values are the same as for PROPDMGEXP.

Note that going forward the updated **sd_edited** dataset was used for all processing.


#### Checking PROPDMGEXP and CROPDMGEXP
PROPDMGEXP and CROPDMGEXP variables were checked to ensure that they contain only H, K, M or B.  

```{r checkExpValues}

        unique(sd_edited$PROPDMGEXP)
        unique(sd_edited$CROPDMGEXP)
```

There are a few values which are not valid.  Since there is nothing in the documentation to indicate what these characters mean,  for the purposes of this analysis they were ignored.  

Total damages for the billions magnitude were summed to get an idea of the figures.
```{r checkExpH}
        # Check billions
                sum(filter(sd_edited, sd_edited$PROPDMGEXP %in% c("B", "b"))$PROPDMG)*1000000000

        # Check hundreds        
                sum(filter(sd_edited, sd_edited$PROPDMGEXP %in% c("H", "h"))$PROPDMG)*100
```

Total damages where the magnitude is in the hundreds is very low compared to the total damages so these figures were ignored.

#### Justification for replacing event names - checking property and crop damage

For both property and crops, invalid events were checked to see how much they contribute to total damages.  

The data required was separated into 2 datasets: **prop** and **crop**.

The following was done to the data:

* Select only the columns that will be used in the analysis: EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP.

* Rows where the damage magnitude is B, M or K, were filtered.  

* The damage magnitude characters were substituted with corresponding numerical values:
        + B or b was replaced with 1 000 000 000
        + M or m was replaced with 1 000 000
        + K or k was replaced with 1 000

* Damages were calculated and added to a new column in each dataset.

* Filter those rows with events which are not in the documentation.

* Find the proportion of the total damages from invalid event names.

```{r percentDamages}

        # Select the appropriate columns
                econtmp <- select(sd_edited, EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)

        # Filter required rows 
                prop <- filter(econtmp, PROPDMGEXP %in% c("B", "b", "M", "m", "K", "k"))
                crop <- filter(econtmp, CROPDMGEXP %in% c("B", "b", "M", "m", "K", "k"))
        
        # Substitute the characters with numerical values
                prop$PROPDMGEXP <- gsub("K", 1000, prop$PROPDMGEXP, ignore.case = TRUE)
                prop$PROPDMGEXP <- gsub("M", 1e+06, prop$PROPDMGEXP, ignore.case = TRUE)
                prop$PROPDMGEXP <- gsub("B", 1e+09, prop$PROPDMGEXP, ignore.case = TRUE)
                prop$PROPDMGEXP <- as.numeric(prop$PROPDMGEXP)
                        
                crop$CROPDMGEXP <- gsub("K", 1000, crop$CROPDMGEXP, ignore.case = TRUE)
                crop$CROPDMGEXP <- gsub("M", 1e+06, crop$CROPDMGEXP, ignore.case = TRUE)
                crop$CROPDMGEXP <- gsub("B", 1e+09, crop$CROPDMGEXP, ignore.case = TRUE)
                crop$CROPDMGEXP <- as.numeric(crop$CROPDMGEXP)        

        # Calculate damages
                prop <- mutate(prop, propDamages = PROPDMG * PROPDMGEXP)
                crop <- mutate(crop, cropDamages = CROPDMG * CROPDMGEXP)

        # Filter rows with invalid events
                invEventsProp <- filter(prop, !(EVTYPE %in% event_types))
                invEventsCrop <- filter(crop, !(EVTYPE %in% event_types))
        
        # Find proportion of total contribution from invalid events
                sum(invEventsProp$propDamages)/sum(prop$propDamages)
                sum(invEventsCrop$cropDamages)/sum(crop$cropDamages)
```

About 18% of property damage and 20% of crop damage can be attributed to invalid events types.  These amounts are large enough to justify replacing some of these names with the documented names.  


#### Description of the selection process for replacing event names

Invalid events in property damage was checked first.

```{r checkInvalidProp}

        # Group by event type and order
                invPropOrdered <- invEventsProp %>%
                        group_by(EVTYPE) %>%
                        summarise (invp_sum = sum(propDamages)) %>%
                        arrange(desc(invp_sum))

        # Checking threshold to use for which events to replace
                lessthan10B_p <- filter(invPropOrdered, invp_sum < 1e+10)
                sum(lessthan10B_p$invp_sum)/sum(prop$propDamages)
                
```

Sum of damages for invalid events which were less than 10 billion contributed about 5% to total property damage.  So it was decided that names would be replaced for those events where property damage was over 10 billion.  

```{r replaceProp}
                morethan10B_p <- filter(invPropOrdered, invp_sum > 1e+10)
                unique(morethan10B_p$EVTYPE)

        # Replace event names
                prop$EVTYPE <- gsub("^HURRICANE .*|^HURRICANE$|^TYPHOON$", "HURRICANE/TYPHOON", 
                                    prop$EVTYPE)
                #prop$EVTYPE <- gsub("^LANDSLIDE$", "DEBRIS FLOW", prop$EVTYPE)
                #prop$EVTYPE <- gsub("^COASTAL FLOODING$", "COASTAL FLOOD", prop$EVTYPE)
                prop$EVTYPE <- gsub("^STORM SURGE.*", "STORM TIDE", prop$EVTYPE)
                #prop$EVTYPE <- gsub("^FLASH FLOOD/.*|FLASH FLOODING", "FLASH FLOOD", prop$EVTYPE)
                #prop$EVTYPE <- gsub("^RIVER FLOOD.*|^FLOOD.*|^MAJOR FLOOD$", "FLOOD", prop$EVTYPE)
                #prop$EVTYPE <- gsub("^SEVERE THUNDERSTORM$|HEAVY RAIN.*", "HEAVY RAIN", prop$EVTYPE)
                #prop$EVTYPE <- gsub("^TORNADOES.*", "TORNADO", prop$EVTYPE)
                #prop$EVTYPE <- gsub("^HAILSTORM$", "HAIL", prop$EVTYPE)
                #prop$EVTYPE <- gsub("^WILDFIRES$", "WILDFIRE", prop$EVTYPE)
                #prop$EVTYPE <- gsub("^HIGH WINDS.*", "HIGH WIND", prop$EVTYPE)

```

A similar exercise was done on invalid event names for crop damages.

```{r checkInvalidCrop}

       # Group by event type and order
                invCropOrdered <- invEventsCrop %>%
                        group_by(EVTYPE) %>%
                        summarise (invc_sum = sum(cropDamages)) %>%
                        arrange(desc(invc_sum))

        # Checking threshold to use for which events to replace
                lessthan1B_c <- filter(invCropOrdered, invc_sum < 1e+9)
                sum(lessthan1B_c$invc_sum)/sum(crop$cropDamages)
```

Sum of damages for invalid events which were less than 1 billion contributed about 4% to total property damage.  So it was decided that names would be replaced for only those invalid events where property damage was over 1 billion.

```{r replaceCrop}
                morethan1B_c <- filter(invCropOrdered, invc_sum >1e+9)
                unique(morethan1B_c$EVTYPE)

        # Replace event names
                crop$EVTYPE <- gsub("^HURRICANE .*|^HURRICANE$", "HURRICANE/TYPHOON", crop$EVTYPE)
                crop$EVTYPE <- gsub("^RIVER FLOOD.*|FLOOD/RAIN.*", "FLOOD", crop$EVTYPE)
                #crop$EVTYPE <- gsub("^FREEZE$|.*FREEZE$", "FLOOD/FREEZE", crop$EVTYPE)
                #crop$EVTYPE <- gsub("^EXCESSIVE WETNESS$", "HEAVY RAIN", crop$EVTYPE)
                
```


# Results
## Weather events most harmful to population health

The edited database, **sd_edited**, was used to create a plot showing the events harmful to the population ie., fatalities and injuries.

First, the injuries were plotted against event type in a barchart.

```{r checkInjSum}

        # Group by Event type, sum injuries for each event and arrange in descending order
                inj_ordered <- sd_edited %>%
                        group_by(EVTYPE) %>%
                        summarise (inj_sum = sum(INJURIES)) %>%
                        arrange(desc(inj_sum))

                head(inj_ordered, 20)
```

The top ten events were sufficient to resonably illustrate the relationship between weather events and injuries.

```{r plotInj}
        # Take the top 10 events
                inj_plotdata <- head(inj_ordered, 10)
        
        # Make EVTYPE an ordered factor so that ggplot does not re-order it
                inj_plotdata$EVTYPE <- factor(inj_plotdata$EVTYPE, levels = inj_plotdata$EVTYPE)

        # Plot injuries
                library(ggplot2)
                injPlot <- ggplot(inj_plotdata, aes(EVTYPE, inj_sum)) +
                        geom_bar(stat = "identity", fill = "#3296FF") +
                        scale_y_continuous(name = "No. of Injuries", breaks=seq(0, 1e+5, 1e+4)) +
                        xlab("Type of Weather Event") +
                        theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

Fatalities were then plotted against event type.  Again the top ten events were used to illustrate the relationship between weather events and fatalities.

```{r plotFatalities}

        # Group by Event type, sum fatalities for each event and arrange in descending order
                fat_ordered <- sd_edited %>%
                        group_by(EVTYPE) %>%
                        summarise (fat_sum = sum(FATALITIES)) %>%
                        arrange(desc(fat_sum))
        
        # Take the top 10 events
                fat_plotdata <- head(fat_ordered, 10)
        
        # Make EVTYPE an ordered factor so that ggplot does not re-order it
                fat_plotdata$EVTYPE <- factor(fat_plotdata$EVTYPE, levels = fat_plotdata$EVTYPE)

        # Plot fatalities
                fatPlot <- ggplot(fat_plotdata, aes(EVTYPE, fat_sum)) +
                        geom_bar(stat = "identity", fill = "#0080C0") +
                        scale_y_continuous(name = "No. of Fatalities", breaks=seq(0, 10000, 1000)) +
                        xlab("Type of Weather Event") +
                        theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

Both charts are displayed in one panel.

```{r dispInjFat}

        library(ggpubr)
        panelInjFat <- ggarrange(injPlot, fatPlot, align = "h")
        annotate_figure(panelInjFat,
                        top = text_grob("Events most harmful to population health", 
                                        color = "navy blue", face = "bold", size = 12),
                        fig.lab = "Figure 1", fig.lab.face = "bold")

```

The barcharts show that tornodoes, thunderstorm winds and excessive heat are the top 3 causes of injuries, while tornadoes, excessive heat and flash flood are the top 3 causes of fatalities.  

Tornadoes, thunderstorm winds, excessive heat, floods, lightning, heat and high wind are the major causes of injuries and fatalities.




## Weather events that have the greatest economic consequences

Property damage was plotted against event type using a barchart.  Again the top ten events were sufficient to illustrate the weather events that caused the greatest damage.

```{r plotPropDamage}

        # Group by event type, sum damages and arrange in descending order
                propOrdered <- prop %>%
                        group_by(EVTYPE) %>%
                        summarise (p_sum = sum(propDamages)) %>%
                        arrange(desc(p_sum))

        # Take the top 10 events
                prop_plotdata <- head(propOrdered, 10)
        
        # Make EVTYPE an ordered factor so that ggplot does not re-order it
                prop_plotdata$EVTYPE <- factor(prop_plotdata$EVTYPE, levels = prop_plotdata$EVTYPE)

        # Plot property damage
                propPlot <- ggplot(prop_plotdata, aes(EVTYPE, p_sum)) +
                        geom_bar(stat = "identity", fill = "#008080") +
                        scale_y_continuous(name = "Amount of Property Damage", 
                                           breaks=seq(0, 1.5e+11, 1e+10)) +
                        xlab("Type of Weather Event") +
                        theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

Crop Damage was then plotted against event type.

```{r plotCropDamage}

        # Group by event type, sum damages and arrange in descending order
                cropOrdered <- crop %>%
                                group_by(EVTYPE) %>%
                                summarise (c_sum = sum(cropDamages)) %>%
                                arrange(desc(c_sum))

        # Take the top 10 events
                crop_plotdata <- head(cropOrdered, 10)
        
        # Make EVTYPE an ordered factor so that ggplot does not re-order it
                crop_plotdata$EVTYPE <- factor(crop_plotdata$EVTYPE, levels = crop_plotdata$EVTYPE)

        # Plot crop damage
                cropPlot <- ggplot(crop_plotdata, aes(EVTYPE, c_sum)) +
                        geom_bar(stat = "identity", fill = "#00CCCC") +
                        scale_y_continuous(name = "Amount of Crop Damage", 
                                           breaks=seq(0, 1.4e+10, 1e+9)) +
                        xlab("Type of Weather Event") +
                        theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

Both plots are displayed in one panel.

```{r displayPropCrop}
        library(ggpubr)
        panelDamage <- ggarrange(propPlot, cropPlot, align = "h")
        annotate_figure(panelDamage,
                        top = text_grob("Events with greatest economic consequences", 
                                        color = "navy blue", face = "bold", size = 12),
                        fig.lab = "Figure 2", fig.lab.face = "bold")
```

The barcharts show that while flood, hurricanes/typhoons and tornadoes are the top 3 causes of property damage, drought, flood and hurricanes/typhoons are the top 3 causes of damage to crops.

Floods, flash floods, hurricanes/typhoons, hail and thunderstorm wind cause substantial damage to both property and crops.

